---
title: "MapBiomas extraction for Trazo field boundaries - v3"
author: "Trazo Fields"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What changed from `extract_mapbiomas_fieldboundaries_chunkedv2.Rmd`

Every change below is backed by a measurement on this machine. Profiles were taken on
Distrito Federal 2019 (50k features), Alagoas 2019 (238k) and Mato Grosso 2024 (1.22M).

### 1. `make_poly_bb_sfc_4326()` is gone — the single biggest win

v2 built one `sfc` rectangle per polygon, in an R loop, purely to test raster overlap.
On Mato Grosso that was **4,017.7 s of a 5,094 s run (79%)**, and it degrades
*superlinearly* — 2.01 -> 2.13 -> **3.28** ms/feature at 50k / 238k / 1.22M, because
`do.call(c, lapply(...))` concatenates a growing `sfc`.

Replaced by a bbox **matrix** plus an arithmetic overlap test: **4,017.7 s -> 14-67 s (60x)**.

> `st_intersects(polys, r_bbox)` is **not** a valid substitute. It throws
> `Loop 0 is not valid: Edge 10 has duplicate vertex with edge 14` on polygonised raster
> boundaries, which carry degenerate edges. That failure is almost certainly why v2 built
> rectangles in the first place. The arithmetic test avoids the geometry engine entirely
> and was verified to give identical answers.

### 2. Filters moved ahead of extraction

v2 extracted MapBiomas for every polygon, and the ipynb Stage G discarded a large share
afterwards. Two filters now run *before* any extraction:

- **`gridcode == 2`** — boundary polygons, dropped by `DROP_GRIDCODE` downstream anyway.
  179,902 of 1,404,888 on Mato Grosso (**12.8%**). Filtered in SQL at read time, so they
  never enter memory.
- **Geography clip** — moved here from ipynb Stage G. Discards **22-71%** depending on
  state (Alagoas -68%, Distrito Federal -71%, Acre -22%); small states lose most because
  110 km S2 tiles overhang their borders.

### 3. Chunking is now unconditional and smaller

v2 chunked only above 1,000,000 features, at 500,000. That is unsafe: `exact_extract(fun =
NULL)` on 1.22M features **OOM'd at 12.6 GB** during `rbindlist`. Chunking is load-bearing,
not a nicety. v3 chunks every layer at `CHUNK_SIZE = 50000`.

| Chunk | Result on Mato Grosso |
|---|---|
| unchunked | 1,076 s extraction, then **OOM at 12.6 GB** |
| 250,000 | **killed** — needs ~19 GB |
| 100,000 | completed, **7.7 GB peak** |
| **50,000 (v3)** | headroom on a machine that had 12.7 GB free |

### 4. Reclassification happens on the pixel table, never the raster

v2 called `terra::classify()` on the whole country raster **once per chunk**. Measured
**135 s vs 2.9 s** cropped on Bolivia — a 46x penalty. `rcl_brazil` is `NULL`, so primary
Brazilian layers escaped it, but **border layers still pay it via neighbour sources**:
Acre extracts against Bolivia and Peru and runs at 15.0 ms/feature against Alagoas' 8.5.

v3 extracts raw values and maps them through a 256-entry lookup vector on the (tiny)
extracted pixel table.

### Two things that were measured and rejected — do not reintroduce

- **crop + built-in C++ summaries** (`exact_extract(r, p, c("mode","count"))`). Measured
  5.7x *faster* on 5,000 contiguous Acre polygons but **2.5x slower** on Mato Grosso. The
  Acre result was a property of spatially compact chunks, not of the method.
- **Hilbert / spatial reordering.** Made extraction **37% slower**, and mean chunk bbox
  area got 17% *larger* (10.23 -> 11.95 deg^2). Polygonisation already emits raster-scan
  order, which is near-optimal for `exact_extract`'s feature-sequential traversal.

### Measurement caveat

This machine is shared with other memory-hungry work (63.8 GB total, ~80% used during
testing). The identical configuration measured **1,781.9 s** and **2,489.7 s** on two runs.
Rankings are reliable; absolute multipliers carry roughly +/-40%. Expect **~2.5-3x** on the
extraction path, multiplied by the 30-75% of features the new filters remove up front.

---

# Settings

```{r settings}
library(terra); library(sf); library(exactextractr); library(data.table)
data.table::setDTthreads(0L)

# Source layers (Stage F output of the ipynb) and destination for finished layers.
fields_dir    <- "D:/WRI/Field Boundaries/Trazo Fields/gpkg_outputs_v2"
out_dir       <- "F:/Trazo Fields v2/field boundaries"
mapbiomas_dir <- "D:/WRI/Field Boundaries/Trazo Fields/extract data/mapbiomas"

# Admin composite built by Stage A of the ipynb (ADM0_NAME / ADM1_NAME / geometry).
boundary_shp  <- "F:/Trazo Fields v2/states and countries/sa_admin_boundaries.shp"

KEEP_QA_COLS              <- TRUE
STRICT_FAIL_ON_NO_MB_YEAR <- FALSE
CAP_TO_LATEST_MB_YEAR     <- TRUE

# Filters applied BEFORE extraction (this is the point of v3).
KEEP_GRIDCODE  <- 1L      # NULL disables the gridcode filter
CLIP_GEOGRAPHY <- TRUE    # moved here from ipynb Stage G

# --- Does nodata count toward mbcov_area? -----------------------------------
# FALSE (v3 default, recommended): nodata cells are excluded from sampled_area.
# TRUE: reproduce v2's behaviour exactly, including its double-counting bug.
#
# v2 summed coverage area across every source whose BOUNDING BOX overlapped a polygon.
# Peru's raster bbox covers all of Acre while containing no Peruvian data there, so those
# cells became class 27 (`others`) and were added to sampled_area -- giving every Acre
# polygon a "coverage area" exactly 2x its real one. Measured on Acre 2019:
#
#   Id 6107:  v2 mbcov 26,119.5 | v3 mbcov 13,059.7 | v2 mbvalid 13,059.7
#
# v2's own mbvalid_area equals the true sampled area, which is what confirms the
# inflation. Ratio was exactly 0.5000 (median and mean) across 14,667 compared features.
#
# CONSEQUENCE: mbcov_area_* in the 64 layers produced by v2 is inflated by roughly the
# number of neighbour rasters whose bbox overlaps that geography. mbmode and
# mbvalid_area are NOT affected (verified: 100% modal agreement, ratio 1.0000).
COUNT_NODATA_IN_COVERAGE <- FALSE

# Chunking: unconditional. 50k leaves headroom on a machine with ~12 GB free;
# 100k peaked at 7.7 GB and 250k was killed outright.
CHUNK_SIZE <- 50000L

# Unset in v2, so GDAL ran on its conservative default. Divide by worker count
# if this is ever parallelised -- each worker carries its own block cache.
Sys.setenv(GDAL_CACHEMAX = "25%")

terra::terraOptions(progress = 0)
OTHERS_VALUE <- 27L
```

---

# Catalogues and helpers (carried over from v2)

```{r country-neighbors}
country_neighbors <- list(
  argentina = c("paraguay", "bolivia", "brazil", "chile", "uruguay"),
  bolivia   = c("brazil", "paraguay", "peru", "argentina", "chile"),
  brazil    = c("bolivia", "paraguay", "peru", "argentina", "uruguay",
                "colombia", "venezuela", "ecuador"),
  chile     = c("argentina", "bolivia", "peru"),
  colombia  = c("ecuador", "peru", "venezuela", "brazil"),
  ecuador   = c("peru", "colombia"),
  paraguay  = c("argentina", "bolivia", "brazil"),
  peru      = c("bolivia", "brazil", "ecuador", "colombia", "chile"),
  uruguay   = c("argentina", "brazil"),
  venezuela = c("colombia", "brazil")
)
```

```{r filename-helpers}
detect_country <- function(x) {
  bn <- tolower(basename(x))
  for (c_ in c("argentina","bolivia","chile","colombia","ecuador",
               "paraguay","peru","uruguay","venezuela"))
    if (grepl(c_, bn)) return(c_)
  if (grepl("brazil|brasil", bn)) return("brazil")
  NA_character_
}

extract_years_from_string <- function(x) {
  hits <- regmatches(x, gregexpr("(?<![0-9])\\d{4}(?![0-9])", x, perl = TRUE))[[1]]
  y <- suppressWarnings(as.integer(hits))
  y[!is.na(y) & y >= 1985 & y <= 2035]
}
parse_mb_file_year <- function(path) {
  y <- extract_years_from_string(basename(path))
  if (!length(y)) NA_integer_ else tail(y, 1)
}
detect_gpkg_year <- parse_mb_file_year

extract_years_from_band_names <- function(nm) {
  y <- suppressWarnings(as.integer(regmatches(nm, regexpr("\\d{4}", nm))))
  y[!is.na(y) & y >= 1985 & y <= 2035]
}
get_mb_available_years <- function(path) {
  fy <- parse_mb_file_year(path)
  if (!is.na(fy)) return(fy)
  r <- rast(path)
  if (nlyr(r) == 1) return(integer(0))
  sort(unique(extract_years_from_band_names(names(r))))
}
```

```{r mb-catalog}
mb_tifs <- list.files(mapbiomas_dir, pattern = "\\.tif$", full.names = TRUE)
if (!length(mb_tifs)) stop("No MapBiomas TIFFs in: ", mapbiomas_dir)

mb_catalog <- data.frame(
  path     = mb_tifs,
  country  = vapply(mb_tifs, detect_country,      character(1)),
  fileyear = vapply(mb_tifs, parse_mb_file_year,  integer(1)),
  exists   = file.exists(mb_tifs), stringsAsFactors = FALSE)
mb_catalog$available_years <- lapply(mb_catalog$path, get_mb_available_years)
mb_catalog$maxyear <- vapply(seq_len(nrow(mb_catalog)), function(k) {
  ay <- mb_catalog$available_years[[k]]
  cand <- c(mb_catalog$fileyear[k], if (length(ay)) ay else NA_integer_)
  if (all(is.na(cand))) NA_integer_ else max(cand, na.rm = TRUE)
}, integer(1))
country_max_year <- tapply(mb_catalog$maxyear, mb_catalog$country, max, na.rm = TRUE)

catalog_has_year <- function(v, y) length(v) > 0 && y %in% v
find_exact_mb_source <- function(cty, yr, cat) {
  rows <- cat[cat$country == cty & cat$exists, , drop = FALSE]
  if (!nrow(rows)) return(NULL)
  for (i in seq_len(nrow(rows)))
    if (!is.na(rows$fileyear[i]) && rows$fileyear[i] == yr ||
        catalog_has_year(rows$available_years[[i]], yr))
      return(list(path = rows$path[i], country = cty, year = yr))
  NULL
}
find_best_mb_source <- function(cty, yr, cat, nbrs) {
  s <- find_exact_mb_source(cty, yr, cat)
  if (!is.null(s)) { s$reason <- "own-country exact-year source"; return(s) }
  for (nb in nbrs[[cty]]) {
    s <- find_exact_mb_source(nb, yr, cat)
    if (!is.null(s)) { s$reason <- paste("neighbor exact-year source:", nb); return(s) }
  }
  NULL
}
```

```{r reclass-tables}
# from -> to per country, carried over from v2. rcl_brazil is NULL (no remapping).
rcl_defs <- list(
  brazil    = NULL, colombia = NULL, venezuela = NULL,
  peru = list(
    from = c( 1, 3, 4, 5, 6,10,11,12,13,14,15,18,21,23,24,25,26,29,30,31,32,33,34,35,40,61,66,68,70,72,27),
    to   = c( 3, 3, 4, 5, 6,12,11,12,12,14,15,18,21,23,24,25,33,29,30,31,32,33,34,35,40,25,12,25,12,41,27)),
  bolivia = list(
    from = c( 1, 3, 4, 5, 6,10,11,12,13,15,18,19,20,21,24,25,27,29,30,31,32,33,34,35,36,39,40, 9,61,66,68,72,81,82),
    to   = c( 3, 3, 4, 5, 6,12,11,12, 3,15,18,41,20,21,24,25,27,29,30,31,32,33,34,35,48,39,40, 9,25,12,25,41,12,11)),
  paraguay = list(
    from = c( 3, 4, 5, 9,11,12,15,18,19,21,24,27,33,39),
    to   = c( 3, 4, 5, 9,11,12,15,18,41,21,24,27,33,39)),
  argentina = list(
    from = c( 3, 4, 5, 9,11,12,15,18,19,20,21,24,25,27,29,30,32,33,34,36,39,41,48,49,50,63,66,73,77),
    to   = c( 3, 4, 5, 9,11,12,15,18,41,20,21,24,25,27,29,30,32,33,34,48,39,41,48,49,50,12, 4,11, 4)),
  ecuador = list(
    from = c( 3, 4, 5, 6, 9,11,12,13,15,18,19,21,24,27,29,30,32,33,34,35,36,39,40,41,68,74,81,82),
    to   = c( 3, 4, 5, 6, 9,11,12,12,15,18,41,21,24,27,29,30,32,33,34,35,48,39,40,41,25,41,12,11)),
  uruguay = list(
    from = c( 3, 9,11,12,15,21,24,27,33,39,41),
    to   = c( 3, 9,11,12,15,21,24,27,33,39,41)),
  chile = list(
    from = c( 3, 4, 5, 9,11,12,15,18,21,23,24,27,29,30,32,33,34,41,48,59,60,61,63,66,67),
    to   = c( 3, 4, 5, 9,11,12,15,18,21,23,24,27,29,30,32,33,34,41,48, 3, 3,25,12, 4, 4))
)

# 256-entry lookup indexed by `value + 1`. Reclassifies any number of pixels with a
# single vectorised subscript -- and never touches the raster.
build_lut <- function(country) {
  d <- rcl_defs[[country]]
  lut <- if (is.null(d)) as.integer(0:255)               # identity (brazil et al.)
         else { l <- rep(OTHERS_VALUE, 256L); l[d$from + 1L] <- as.integer(d$to); l }
  # Class 0 is nodata. Mapping it to NA drops it from sampled_area (v3 default);
  # leaving it in reproduces v2, where 0 became `others` and inflated the column.
  # Either way it is excluded from validclass_area, which filters val > 0 & val != 27.
  if (!COUNT_NODATA_IN_COVERAGE) lut[1L] <- NA_integer_
  as.integer(lut)
}
LUT_CACHE <- new.env(parent = emptyenv())
get_lut <- function(country) {
  if (is.null(LUT_CACHE[[country]])) LUT_CACHE[[country]] <- build_lut(country)
  LUT_CACHE[[country]]
}
```

---

# v3 core: fast bounding boxes, geography lookup, chunked extraction

```{r fast-bbox}
# Feature bounding boxes as a plain matrix (xmin ymin xmax ymax per row).
# ~156x faster than v2's sfc-rectangle loop: no geometry objects are constructed.
feature_bboxes <- function(x) {
  t(vapply(sf::st_geometry(x), function(g) as.numeric(sf::st_bbox(g)), numeric(4)))
}

# Overlap against a raster extent, by arithmetic. Identical to v2's
# st_intersects(bbox_rectangles, raster_bbox), and immune to invalid geometry.
bbox_overlaps_raster <- function(bb, r) {
  rb <- as.numeric(sf::st_bbox(terra::ext(r)))
  bb[, 3] >= rb[1] & bb[, 1] <= rb[3] & bb[, 4] >= rb[2] & bb[, 2] <= rb[4]
}
```

```{r geography-lookup}
# Accent/case-insensitive key, matching the ipynb's norm() so layer names built from
# folder names ("Brazil_Mato_Grosso_Do_Sul_2024") match GADM's "Mato Grosso do Sul".
norm_key <- function(s) {
  s <- iconv(as.character(s), to = "ASCII//TRANSLIT")
  s[is.na(s)] <- ""
  s <- gsub("[_]+", " ", s)
  tolower(trimws(gsub("\\s+", " ", s)))
}

admin <- st_read(boundary_shp, quiet = TRUE)
if (is.na(st_crs(admin))) admin <- st_set_crs(admin, 4326)
admin <- st_transform(admin, 4326)
admin$k <- paste(norm_key(admin$ADM0_NAME), norm_key(admin$ADM1_NAME), sep = "|")
cat("admin units:", nrow(admin), "\n")

STATE_LEVEL <- c("Brazil", "Argentina")

# "Brazil_Mato_Grosso_2024" -> list(adm0="Brazil", adm1="Mato_Grosso", year=2024)
parse_layer_name <- function(layer) {
  m <- regmatches(layer, regexec("^(.+)_(\\d{4})$", layer))[[1]]
  if (length(m) != 3) return(NULL)
  region <- m[2]; year <- as.integer(m[3])
  for (c_ in STATE_LEVEL)
    if (startsWith(region, paste0(c_, "_")))
      return(list(adm0 = c_, adm1 = substring(region, nchar(c_) + 2L), year = year))
  list(adm0 = region, adm1 = "", year = year)
}

layer_geometry <- function(layer) {
  p <- parse_layer_name(layer)
  if (is.null(p)) return(NULL)
  k <- paste(norm_key(p$adm0), norm_key(p$adm1), sep = "|")
  i <- which(admin$k == k)
  if (!length(i)) return(NULL)
  st_geometry(admin)[i[1]]
}
```

```{r extraction-engine}
# Area-weighted modal class per polygon. Ties break toward the smaller class, matching
# v2's `setorder(cls, pid, -a, val)`. Measured: zero exact ties in 20,000 polygons --
# coverage_area is a continuous float, so ties are effectively measure-zero.
aggregate_chunk <- function(d, n_chunk) {
  out <- data.table(k = seq_len(n_chunk), sampled_area = 0,
                    validclass_area = 0, mb_mode = NA_real_)
  if (!nrow(d)) return(out)
  d <- d[!is.na(val) & !is.na(area)]
  if (!nrow(d)) return(out)

  out[d[, .(s = sum(area)), by = k], sampled_area := i.s, on = "k"]

  vv <- d[val > 0 & val != OTHERS_VALUE]
  if (!nrow(vv)) return(out)
  out[vv[, .(v = sum(area)), by = k], validclass_area := i.v, on = "k"]

  cls <- vv[, .(a = sum(area)), by = .(k, val)]
  setorder(cls, k, -a, val)
  out[cls[, .(m = as.numeric(val[1])), by = k], mb_mode := i.m, on = "k"]
  out[]
}

# Extract one chunk against every source (primary + neighbours), pooling pixels.
# Each source is reclassified through ITS OWN LUT on the pixel table -- the raster is
# never modified, which is what removes v2's per-chunk whole-country classify().
extract_chunk <- function(polys_chunk, bb_chunk, sources) {
  n <- nrow(polys_chunk)
  parts <- list()
  for (s in sources) {
    r  <- rast(s$path)
    ov <- bbox_overlaps_raster(bb_chunk, r)
    if (!any(ov)) next
    pix <- exact_extract(r, polys_chunk[ov, ], fun = NULL,
                         coverage_area = TRUE, progress = FALSE)
    d <- rbindlist(pix, idcol = "j", use.names = TRUE, fill = TRUE)
    if (!nrow(d)) next
    setnames(d, c("value", "coverage_area"), c("val", "area"), skip_absent = TRUE)
    d[, val := get_lut(s$country)[as.integer(val) + 1L]]
    d[, k := which(ov)[j]][, j := NULL]
    parts[[length(parts) + 1L]] <- d
    rm(pix, d)
  }
  if (!length(parts)) {
    return(data.table(k = seq_len(n), sampled_area = 0,
                      validclass_area = 0, mb_mode = NA_real_))
  }
  aggregate_chunk(rbindlist(parts, use.names = TRUE, fill = TRUE), n)
}
```

---

# Main loop

```{r main-loop}
srcs <- list.files(fields_dir, pattern = "\\.gpkg$", full.names = TRUE)
srcs <- srcs[!grepl("__part[0-9]{3}\\.gpkg$", srcs)]
cat("source layers:", length(srcs), "\n\n")

log_rows <- list()

for (src in srcs) {
  layer    <- tools::file_path_sans_ext(basename(src))
  out_gpkg <- file.path(out_dir, paste0(layer, ".gpkg"))
  t0 <- proc.time()

  cat("==============================================================\n")
  cat("Layer:", layer, "\n")

  # Skip completed layers (an mbmode column means v2 or v3 already finished it).
  if (file.exists(out_gpkg)) {
    cols <- tryCatch(names(st_read(out_gpkg,
              query = sprintf('SELECT * FROM "%s" LIMIT 0', layer), quiet = TRUE)),
            error = function(e) character(0))
    if (any(grepl("^mbmode", cols))) { cat("  SKIP (already has mbmode)\n\n"); next }
  }

  country_key <- detect_country(src)
  file_yr     <- detect_gpkg_year(src)
  if (is.na(country_key)) { cat("  cannot detect country - skip\n\n"); next }
  if (is.na(file_yr))     { cat("  cannot detect year - skip\n\n");    next }

  # ---- 1. read, with gridcode filtered AT SOURCE -------------------------
  q <- sprintf('SELECT * FROM "%s"', layer)
  if (!is.null(KEEP_GRIDCODE)) q <- paste0(q, sprintf(" WHERE gridcode = %d", KEEP_GRIDCODE))
  polys <- tryCatch(st_read(src, query = q, quiet = TRUE),
                    error = function(e) { cat("  ERROR reading:", conditionMessage(e), "\n"); NULL })
  if (is.null(polys) || !nrow(polys)) { cat("  0 features - skip\n\n"); next }
  n_gc <- nrow(polys)
  cat("  read (gridcode ==", KEEP_GRIDCODE, "):", format(n_gc, big.mark = ","), "\n")

  if (is.na(st_crs(polys))) polys <- st_set_crs(polys, 4326)
  if (st_crs(polys) != st_crs(4326)) polys <- st_transform(polys, 4326)

  # ---- 2. geography clip, BEFORE extraction ------------------------------
  # s2 MUST be off here. With s2 on this throws
  #   "Loop 0 is not valid: Edge 24 has duplicate vertex with edge 28"
  # because polygonised raster boundaries carry degenerate edges. Planar is also the
  # correct semantics to match: the ipynb Stage G used geopandas, i.e. planar GEOS.
  if (CLIP_GEOGRAPHY) {
    geom <- layer_geometry(layer)
    if (is.null(geom)) {
      cat("  WARNING: no admin geometry for", layer, "- clip skipped\n")
    } else {
      s2_was <- sf_use_s2()
      suppressMessages(sf_use_s2(FALSE))
      keep <- tryCatch(lengths(st_intersects(polys, geom)) > 0,
                       error = function(e) {
                         cat("  clip FAILED:", conditionMessage(e), "- keeping all\n")
                         rep(TRUE, nrow(polys))
                       })
      suppressMessages(sf_use_s2(s2_was))
      polys <- polys[keep, ]
      cat("  after geography clip:", format(nrow(polys), big.mark = ","),
          sprintf("(-%.1f%%)\n", 100 * (1 - nrow(polys) / n_gc)))
    }
  }
  n <- nrow(polys)
  if (!n) { cat("  0 features after clip - skip\n\n"); next }

  # ---- 3. year cap + output column names --------------------------------
  cmax  <- suppressWarnings(country_max_year[[country_key]])
  mb_yr <- if (CAP_TO_LATEST_MB_YEAR && !is.null(cmax) && !is.na(cmax) && file_yr > cmax) {
             cat("  capping MapBiomas year", file_yr, "->", cmax, "\n"); cmax
           } else file_yr
  out_col <- paste0("mbmode",        substr(file_yr, 3, 4))
  cov_col <- paste0("mbcov_area_",   file_yr)
  val_col <- paste0("mbvalid_area_", file_yr)

  srcmb <- find_best_mb_source(country_key, mb_yr, mb_catalog, country_neighbors)
  if (is.null(srcmb)) {
    if (STRICT_FAIL_ON_NO_MB_YEAR) stop("No MapBiomas source for ", country_key, " ", mb_yr)
    cat("  no MapBiomas source - writing NA columns\n")
    polys[[out_col]] <- NA_real_; polys[[cov_col]] <- 0; polys[[val_col]] <- 0
    st_write(polys, out_gpkg, layer = layer, delete_dsn = TRUE, quiet = TRUE)
    cat("\n"); next
  }

  sources <- c(list(list(path = srcmb$path, country = srcmb$country)),
               Filter(Negate(is.null), lapply(country_neighbors[[country_key]], function(nb) {
                 s <- find_exact_mb_source(nb, mb_yr, mb_catalog)
                 if (is.null(s)) NULL else list(path = s$path, country = nb)
               })))
  cat("  sources:", paste(vapply(sources, `[[`, character(1), "country"), collapse = ", "), "\n")

  # ---- 4. bbox matrix once, then chunked extraction -----------------------
  tb <- proc.time()
  BB <- feature_bboxes(polys)
  cat(sprintf("  bbox matrix: %.1f s (%.3f ms/feat)\n",
              (proc.time() - tb)[["elapsed"]], 1000 * (proc.time() - tb)[["elapsed"]] / n))

  mode_v <- rep(NA_real_, n); cov_v <- rep(0, n); val_v <- rep(0, n)
  brks <- seq(1L, n, by = CHUNK_SIZE)
  te <- proc.time()
  for (ci in seq_along(brks)) {
    sel <- brks[ci]:min(brks[ci] + CHUNK_SIZE - 1L, n)
    res <- extract_chunk(polys[sel, ], BB[sel, , drop = FALSE], sources)
    mode_v[sel] <- res$mb_mode; cov_v[sel] <- res$sampled_area; val_v[sel] <- res$validclass_area
    rm(res); invisible(gc(verbose = FALSE))
    cat(sprintf("    chunk %d/%d  %.0f s cum\n", ci, length(brks),
                (proc.time() - te)[["elapsed"]]))
  }
  t_ext <- (proc.time() - te)[["elapsed"]]

  # ---- 5. write ----------------------------------------------------------
  polys[[out_col]] <- mode_v
  if (KEEP_QA_COLS) { polys[[cov_col]] <- cov_v; polys[[val_col]] <- val_v }
  st_write(polys, out_gpkg, layer = layer, delete_dsn = TRUE, quiet = TRUE)

  n_out <- as.integer(st_read(out_gpkg,
            query = sprintf('SELECT COUNT(*) AS n FROM "%s"', layer), quiet = TRUE)$n)
  el <- (proc.time() - t0)[["elapsed"]]
  cat(sprintf("  wrote %s features %s | %.0f s total (%.3f ms/feat extraction)\n\n",
              format(n_out, big.mark = ","),
              if (identical(n_out, n)) "(verified)" else "(*** COUNT MISMATCH ***)",
              el, 1000 * t_ext / n))

  log_rows[[length(log_rows) + 1L]] <- data.table(
    layer = layer, n_gridcode = n_gc, n_after_clip = n, n_written = n_out,
    total_s = el, extract_s = t_ext, ms_per_feat = 1000 * t_ext / n)

  rm(polys, BB, mode_v, cov_v, val_v); invisible(gc(verbose = FALSE))
}

if (length(log_rows)) {
  run_log <- rbindlist(log_rows)
  print(run_log)
  fwrite(run_log, file.path(dirname(out_dir), "v3_run_log.csv"))
}
```

---

# Verification against v2

Run this on a layer v2 already produced. It re-extracts with v3 and compares row by row.
**Do this before trusting a production run** — the `"sum"` vs `"count"` bug found during
development produced a silently wrong area column while modal class stayed 100% correct,
so agreement on one field proves nothing about the others.

```{r verify, eval=FALSE}
VERIFY_LAYER <- "Brazil_Acre_2019"
v2_path <- file.path(out_dir, paste0(VERIFY_LAYER, ".gpkg"))

v2 <- st_read(v2_path, quiet = TRUE)
mode_col <- grep("^mbmode", names(v2), value = TRUE)[1]

src <- file.path(fields_dir, paste0(VERIFY_LAYER, ".gpkg"))
q <- sprintf('SELECT * FROM "%s" WHERE gridcode = %d', VERIFY_LAYER, KEEP_GRIDCODE)
p <- st_transform(st_read(src, query = q, quiet = TRUE), 4326)
g <- layer_geometry(VERIFY_LAYER)
if (!is.null(g)) {
  suppressMessages(sf_use_s2(FALSE))
  p <- p[lengths(st_intersects(p, g)) > 0, ]
  suppressMessages(sf_use_s2(TRUE))
}

# The R clip does not reproduce the ipynb's Python clip EXACTLY: measured 141,562 vs
# 141,560 on Acre 2019, a +2 / 0.0014% difference. Both use planar GEOS `intersects`,
# so the discrepancy is boundary-touching polygons resolving differently between
# geopandas and sf. Tolerate a small difference; a large one means the wrong geometry
# was matched (check parse_layer_name / norm_key against the admin table).
cat("v2 rows:", nrow(v2), " v3 rows:", nrow(p),
    sprintf(" (diff %+d, %.4f%%)\n", nrow(p) - nrow(v2),
            100 * abs(nrow(p) - nrow(v2)) / nrow(v2)))
if (abs(nrow(p) - nrow(v2)) / nrow(v2) > 0.001)
  stop("clip differs from ipynb Stage G by more than 0.1% - investigate before proceeding")

# Row-wise comparison below requires identical row sets. When counts differ by a
# handful of boundary polygons, compare on the intersection by geometry id instead.
if (nrow(p) != nrow(v2)) {
  cat("NOTE: row counts differ; comparing on the common `Id` set only.\n")
  common <- intersect(v2$Id, p$Id)
  v2 <- v2[match(common, v2$Id), ]
  p  <- p [match(common, p$Id),  ]
  cat("comparing", length(common), "common features\n")
}

cy <- detect_gpkg_year(src); ck <- detect_country(src)
sm <- find_best_mb_source(ck, cy, mb_catalog, country_neighbors)
sources <- c(list(list(path = sm$path, country = sm$country)),
             Filter(Negate(is.null), lapply(country_neighbors[[ck]], function(nb) {
               s <- find_exact_mb_source(nb, cy, mb_catalog)
               if (is.null(s)) NULL else list(path = s$path, country = nb)
             })))

BB <- feature_bboxes(p); n <- nrow(p)
m3 <- rep(NA_real_, n); c3 <- rep(0, n)
for (lo in seq(1L, n, by = CHUNK_SIZE)) {
  sel <- lo:min(lo + CHUNK_SIZE - 1L, n)
  r_  <- extract_chunk(p[sel, ], BB[sel, , drop = FALSE], sources)
  m3[sel] <- r_$mb_mode; c3[sel] <- r_$sampled_area
}

agree <- mean((is.na(v2[[mode_col]]) & is.na(m3)) |
              (!is.na(v2[[mode_col]]) & !is.na(m3) & v2[[mode_col]] == m3))
cat(sprintf("mb_mode agreement : %.4f%%\n", 100 * agree))

cov_col <- grep("^mbcov_area_", names(v2), value = TRUE)[1]
if (!is.na(cov_col)) {
  rel <- abs(v2[[cov_col]] - c3) / pmax(v2[[cov_col]], 1e-9)
  cat(sprintf("max rel. area diff: %.2e\n", max(rel, na.rm = TRUE)))
}
cat("\nRequire ~100%% mode agreement and area diff <= ~1e-7 before running production.\n")
```

```{r session-info}
sessionInfo()
```
